Eye Fitting Straight Lines in the Modern Era

### 2021-09-19 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
Some edits to intro/abstract

abstract: |
  Fitting lines by eye through a set of points has been explored since the 20th century. Common methods of fitting trends by eye involve maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points. In 2015, the New York Times introduced an interactive feature, called 'You Draw It'. Readers are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality. The New York Times team utilizes Data Driven Documents (D3) that allows readers to predict these metrics by drawing a line on their computer screen with their computer mouse. In my research, I established ‘You Draw It’ as a method for graphical testing by adapting the New York Times feature. I recruited participants via crowdsourcing websites and replicated the study found in Eye Fitting Straight Lines (Mosteller et al., 1981). Participants were directed to an RShiny application link and shown points following a linear trend and asked to draw a line through the data points using their computer mouse; task plots were generated using the r2d3 package in R statistical software. Results from my study were consistent with those found in the previous study; when shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the least-squares regression line. This trend was most prominent when shown data simulated with larger variances. The reproducibility of these results serves as evidence of the reliability of the you draw it method. Future work is necessary to implement the 'You Draw It' tool as a method of testing graphics. [200 word limit]

@@

Data visualization has become a central tool in modern data science.
@unwin2020data defines data visualization as the art of graphic displays to show data. 
Graphics illustrate the relationship between two sets of information or variables. 
Graphical displays are useful for data cleaning, exploratory data analysis, identifying trends, evaluating model output, and presenting results.
When creating a graphic, we must consider the design choices most effective for conveying the intended result.
In order to evaluate the effectiveness of a graphic, we must understand cognitive perception and biases related to different types of data structures. 
Efforts in the field of graphics have developed graphical testing tools and methods such as the lineup protocol [@buja2009statistical] to provide a framework for inferential testing.
Through experimentation, methods such as the lineup protocol allow researchers to conduct studies geared at understanding human ability to conduct tasks related to the perception of statistical charts such as differentiation, prediction, estimation, and extrapolation.

>>>>

abstract: |
  \svp{How do statistical regression results compare to intuitive, visually fitted results?} Fitting lines by eye through a set of points has been explored since the 20th century. Common methods of fitting trends by eye involve maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points. In 2015, the New York Times introduced an interactive feature, called 'You Draw It'\svp{, where r}eaders are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality. \svp{In this paper, we validate} ‘You Draw It’ as a method for graphical testing, \svp{comparing results to the less technological method utilized in \citet{mosteller1981eye} and extending that study with formal statistical analysis methods.} Results were consistent with those found in the previous study; when shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the least-squares regression line. This trend was most prominent when shown data simulated with larger variances. \svp{This study reinforces the differences between intuitive visual model fitting and statistical model fitting, providing information about human perception as it relates to the use of statistical graphics.}

@@

Data visualization has become a central tool in modern data science.\svp{XXX this is a bit of a lame intro... we might want to spice this up later, but not critical now}
@unwin2020data defines data visualization as the art of \svp{using} graphic displays to show data. \svp{XXX how does this relate - need to make it flow a bit better. Add additional clauses or connect this to the next thought somehow.}
Graphics illustrate the relationship between two sets of information or variables. 
Graphical displays are useful for data cleaning, exploratory data analysis, identifying trends, evaluating model output, and presenting results.\svp{Right now this reads like bullet points rather than a narrative.}

When creating a graphic, we must consider the design choices most effective for conveying the intended result. \svp{For instance, we must decide to highlight the relationship between two variables in a scatterplot by including a trend line, or adding color to highlight clustering \citep{vanderplas2017clusters}.}
These design choices require that we understand the perception and visual biases that come into play when creating graphics, and as graphics are evaluated visually, we must use human testing to ground our understanding in empiricism.}
<!-- In order to evaluate the effectiveness of a graphic, we must understand cognitive perception and biases related to different types of data structures.  -->
Efforts in the field of graphics have developed graphical testing tools and methods such as the lineup protocol [@buja2009statistical] to provide a framework for inferential testing.
Through experimentation, methods such as the lineup protocol allow researchers to conduct studies geared at understanding human ability to conduct tasks related to the perception of statistical charts such as differentiation, prediction, estimation, and extrapolation.
